---
title: "duckplyr: A trustworthy drop-in accelerator for dplyr workflows"
author: "R Stats Club"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    df_print: paged
---

## Motivation and Goals

- Explain what `duckplyr` is and how it integrates with `dplyr`.
- Show how `library(duckplyr)` effectively changes `dplyr` behavior (without rewriting code).
- Argue why `duckplyr` is more trustworthy than `dtplyr`, using a concrete join example.
- Walk through a mini end-to-end workflow using `read_csv_duckdb()`.
- Briefly introduce lazy execution, query optimization, and prudence.
- Conclude with practical guidance on when to use `duckplyr`.

We assume:

- Some familiarity with `dplyr`.
- Basic awareness of DuckDB (but not deep internals).

## What Is duckplyr?

`duckplyr` is an integration between DuckDB and `dplyr`.

It lets you write ordinary `dplyr` pipelines while:

- Building a lazy query plan.
- Letting DuckDB optimize and execute the query.
- Only materializing results into R when needed.

This can lead to:

- Faster execution.
- Lower memory usage.
- The ability to work directly on files (CSV / Parquet).

## Attaching duckplyr: What Actually Changes

```{r}
library(tidyverse)
library(duckplyr)
```

When you attach `duckplyr`, you'll see a startup message indicating that some `dplyr` methods are being overwritten.

Conceptually:

- You still call `left_join()`, `mutate()`, `summarise()`, etc.
- `duckplyr` inserts itself so that supported verbs can be executed by DuckDB.
- Unsupported operations fall back to `dplyr`.
- Under the hood this uses S3 method dispatch and class-based behavior — it behaves like a drop-in replacement.

## Restoring Original dplyr Behavior

```{r}
methods_restore()
```

## Fallback Behavior and Diagnostics

`duckplyr` only runs code in DuckDB when it can do so safely. If a feature is unsupported, `duckplyr`:

- Falls back to `dplyr`.
- Can explicitly tell you why.

Enable fallback diagnostics:

```{r}
fallback_config(info = TRUE)
```

This is critical for understanding what is (and is not) accelerated.

## Why Trust duckplyr More Than dtplyr?

In my experience, `dtplyr` has silently mistranslated `dplyr` code, whereas
`duckplyr` seems to more carefully identify when an operation is not fully
supported, and explicitly lets you know when this happens. In real code, I
tried out a fairly obscure parameter `multiple = "any"` in `left_join()`;
here we'll show how `duckplyr` handles this case compared to `dtplyr`.

### Example: `left_join(multiple = "any")`

Setup: two small tibbles.

```{r}

keys <- tibble(
    key = c("A", "B", "C")
)

anno <- tibble(
    key = c("A","A","A","B","B","C","C","C","C"),
    gene = c("G1","G2","G3","G4","G5","G6","G7","G8","G9"),
    cluster = c(1,1,1,2,2,3,3,3,3)
)

keys
anno
```

- `keys` has unique keys.
- `anno` has multiple rows per key.
- Each key maps to exactly one `cluster`.

#### dplyr semantics (`multiple = "any"`)

```{r}
res_dplyr_any <- left_join(
    keys,
    anno,
    by = "key",
    multiple = "any"
)

res_dplyr_any
nrow(res_dplyr_any)
```

Expected behavior:

- Exactly one row per key.
- One arbitrary matching row chosen per key.
- Total rows = `nrow(keys)`.

#### dtplyr mistranslation

```{r}
library(dtplyr)

lazy_keys <- lazy_dt(keys)
lazy_anno <- lazy_dt(anno)

res_dtplyr_any <- left_join(
        lazy_keys,
        lazy_anno,
        by = "key",
        multiple = "any"
    ) |>
    as_tibble()

res_dtplyr_any
nrow(res_dtplyr_any)
```

The logic behind `multiple = "any"` was completely ignored, and worse, the difference
in behavior was silent. This particular example is still an open GitHub issue
[here](https://github.com/tidyverse/dtplyr/issues/488).

#### duckplyr: correct result with explicit fallback

Ideally, make sure fallback diagnostics are enabled, then run the same join. Note that
we earlier called `methods_restore()`, and so must re-enable the `duckplyr` methods first:

```{r}
methods_overwrite()

res_duckplyr_any <- left_join(
    keys,
    anno,
    by = "key",
    multiple = "any"
)

res_duckplyr_any
nrow(res_duckplyr_any)
```

Note the clear warning message that DuckDB cannot process the `multiple = "any"` argument
and is falling back to `dplyr` to still produce identical results.

## Mini End-to-End Workflow with `read_csv_duckdb()`

As an example, we'll use a tiny dataset, but of course the real benefits come
when working with much larger datasets.

### Create a small CSV

```{r}
tmp_csv <- tempfile(fileext = ".csv")

demo_tbl <- tibble(
    sample_id = rep(c("S1", "S2"), each = 6),
    gene = rep(paste0("G", 1:6), times = 2),
    count = c(10, 0, 3, 7, 1, 0, 4, 9, 0, 2, 8, 1),
    cluster = rep(c("C1","C2","C3"), times = 4)
)

write_csv(demo_tbl, tmp_csv)
```

### Run a dplyr pipeline

Unfortunately in this case the syntax must be changed subtly to work with `duckplyr` (note the
`.by` argument in `summarise()` replacing `group_by()`).

```{r}
pipeline <- read_csv_duckdb(tmp_csv) |>
    filter(count > 0) |>
    summarise(
        .by = c("sample_id", "cluster"),
        n_genes = n(),
        total = sum(count)
    ) |>
    arrange(desc(total))
```

At this stage:

- The computation is represented as a query plan.
- No data has been materialized or even read into R yet

### Explicitly materialize results

```{r}
result <- pipeline |> collect()
result
```

This is where data is actually pulled into R and manipulated.

## A Brief Note on Prudence

Prudence controls how eagerly `duckplyr` materializes data into R.

- `lavish`: behave more like a regular tibble.
- `thrifty` / `stingy`: require more explicit `collect()`.

To always get the expected behavior in `dplyr`, you'll have to explicitly request
lavish prudence.

```{r, eval = FALSE}
df_big <- read_csv_duckdb("big.csv", prudence_level = "lavish")
```

Why this matters:

- Automatic materialization can accidentally pull large data into memory.
- Stricter prudence protects you, but feels less like base `dplyr`.
- Choose based on data size and workflow.

## Summary and Takeaways

- `duckplyr` is designed as a drop-in accelerator for `dplyr` workflows.
- It prioritizes correctness and transparency.
- Unsupported operations fall back loudly, not silently.
- Lazy execution + DuckDB optimization can yield noticeable performance gains.
- Prudence lets you balance convenience vs safety.
