---
title: "Introduction to dtplyr"
author: 
  - name: Nicholas J. Eagles
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  %\VignetteIndexEntry{Introduction to dtplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

`dplyr` provides an excellent, expressive language for manipulating data frames
in R. After a slight learning curve, it arguably produces more readable code
while enabling complex, powerful operations. However, a main drawback is that it
wasn't designed to prioritize memory efficiency or speed. `data.table` was
developed as a performant alternative, but has a completely different syntax
that can feel unnatural to those used to base R or `dplyr`.

`dtplyr` bridges this gap, allowing users to write `dplyr` code that *mostly*
gets executed as `data.table` code under the hood. While it can somewhat be
treated as a black box, I find that understanding the logic underlying memory
allocation for various operations is useful to write the most performant code
while avoiding unexpected behavior.

# Understanding Memory Allocation in R

First, let's talk about memory allocation in base R, which feels quite similar
to how `dplyr` works. In particular, both protect the user from accidentally
modifying an existing object during a series of modifications, often at the
expense of memory efficiency.

## R's Default Behavior: Copy on Modify

```{r 'copy_on_modify'}
library(dplyr)
library(data.table)
library(dtplyr)

a = tibble(x = 1:3)

#   At this point, a second tibble is not created. 'b' just points to the same
#   location in memory as 'a'; it's a reference, not copy
b = a

#   However, the second we modify 'b', R actually creates a full copy of 'a',
#   then adds a column to the new tibble
b$y = 4:6

#   Note that 'a' remains unchanged
a
```

This behavior applies for almost all ordinary objects in base R, like vectors,
lists, and data frames!

## `dplyr`'s behavior

Somewhat similarly, adding a column to a data frame with `dplyr` creates a full
copy of the starting data frame before adding the new column:

```{r 'dplyr_mutate'}
a = tibble(x = 1:3)

#   'a' is fully copied, then the new column is added to 'b', the copy
b = mutate(a, y = 4:6)
```

Actually, every `dplyr` verb creates a copy of at least some of the starting
data, even, for example, `select()`, `filter()`, and `arrange()`.

## `data.table`'s Behavior (Certain Operations)

In contrast, `data.table` tries to avoid data copies as much as possible and
instead use references to the original object. Let's again consider the example
of adding a column to an existing data frame:

```{r 'data_table_mutate'}
a = data.table(x = 1:3)

#   Use the ':=' operator to add a column by reference. The existing data
#   stays where it is (the 'x' column), and the new column 'y' is added to 'a'
a[, y := 4:6]
```

Other examples of operations that modify a reference include functions that
start with `set`. For example, `setnames()` changes colnames by reference
(shockingly, `colnames<-` does not-- some underlying data is copied!):

```{r 'data_table_setnames'}
setnames(a, c("x_new", "y_new"))

a
```

Another handy function is `setDT()`, which can take a data-frame-like object
(e.g. a `tibble` or `data.frame`) and convert it to a `data.table` by reference.
If you use `data.table()` to cast the starting object, it gets copied!

# Introducing `dtplyr`

`dtplyr` works by introducing a class informally called a "lazy data table".
Code can be written with these `lazy_dt()` objects as if they were ordinary
data tables, but any `dplyr` verbs are not actually executed until the object
is converted to the ordinary type (e.g. `as.data.frame()` or `as_tibble()`).
The code is executed as `data.table` code, often with performance gains relative
to the written `dplyr` code.

## Basic Example

Let's grab a copy of `mtcars` and use a `lazy_dt()` to translate some `dplyr`
operations into more efficient `data.table` code.

```{r 'dtplyr_lazy_dt'}
a = data.table(mtcars)

#   Need to convert our input data to a lazy data table
a = lazy_dt(a)

#   Lazily track the following dplyr operations. They don't get performed yet!
a = a |>
    group_by(cyl) |>
    summarize(mean_mpg = mean(mpg)) |>
    arrange(cyl)

#   We can check the data.table code that will be executed upon conversion to
#   a tibble or similar
show_query(a)

#   Now perform the conversion, and execute the dplyr operations as data.table
#   code
a = as_tibble(a)
a
```

## Efficiency Caveats

`dtplyr` does not always perform the most efficient `data.table` translations of
`dplyr` code. The major exception is that `dplyr::mutate()` still creates a copy
of the starting data frame (by default) through `dtplyr`. This is intentional, as
it ensures `dplyr` code in `dtplyr` behaves the same as usual.

Let's turn this copying behavior off, with a warning about side effects:

```{r 'dtplyr_mutate_no_copy'}
a = data.table(mtcars)

#   Explicitly specify that we're allowed to modify the original object
b = lazy_dt(a, immutable = FALSE)

#   Add some lazy operations
b = b |>
    mutate(new_col = "some_constant") |> # memory-efficient mutate
    group_by(cyl) |>
    summarize(mean_mpg = mean(mpg)) |>
    arrange(cyl)

b = as_tibble(b)

#   But wait, 'a' has a new column 'new_col'
head(a)
```
